---
title: "MLE_Methods"
output:
  word_document: default
  html_document: default
  pdf_document: default
date: "2026-04-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Maximum Likelihood Estimation

MLE can be used to determine the best parameters for each of the models for the best fit of the data. The Beta-Poisson model is considered more complex because of its additional parameter. However, the model may not be that much better than a simpler version when considering factors such as overfitting and AIC.

Comparing AIC for three models: Poisson, Negative Binomial, and Beta-Poisson
The standard Poisson model had the highest AIC (310.11), indicating that the data is overdispersed (due to superspreaders)."Furthermore, the results show that even with the extra parameter, the Beta-Poisson model had a higher AIC (308.96) than the negative binomial model, and ultimately showed that negative binomial (306.96) is the more parsimonious model.
```{r}
transmissiondata <- read.csv("transmission_pairs.csv")
data_vec <- aggregate(infectee.case ~ infector.case,data = transmissiondata,FUN = length)
colnames(data_vec)<- c("infector.case", "secondary_cases")
```

```{r}
library(MASS)

y <- data_vec$secondary_cases

#Poisson Model
fit_pois <- glm(y ~ 1, family = "poisson")
aic_pois <- AIC(fit_pois)

#Negative Binomial Model
fit_nb <- glm.nb(y ~ 1)
aic_nb <- AIC(fit_nb)

cat("Poisson AIC:", aic_pois, "\n")
cat("Neg Binomial AIC:", aic_nb, "\n")
```
```{r}
library(hypergeo)

beta_pois_nll <- function(params, data) {
  lambda <- params[1]
  phi    <- params[2]
  N      <- params[3]
  
  if (lambda <= 0 || phi <= 0 || N <= lambda) return(1e10)
  
  a1 <- lambda * phi
  a2 <- (N - lambda) * phi
  
  log_probs <- sapply(data, function(x) {
    log_const <- x * log(N) - lfactorial(x)
    
    log_beta_ratio <- (lbeta(a1 + x, a2) - lbeta(a1, a2))
    f1f1 <- Re(genhypergeo(U = a1 + x, L = a1 + a2 + x, z = -N))
    
    return(log_const + log_beta_ratio + log(f1f1 + 1e-10))
  })
  
  return(-sum(log_probs))
}

#Beta-Poisson Model
fit_bp <- optim(par = c(mean(y), 1, max(y) + 1), 
                fn = beta_pois_nll, 
                data = y,
                method = "Nelder-Mead")

aic_bp <- 2 * 3 + 2 * fit_bp$value
cat("Beta-Poisson AIC:", aic_bp, "\n")
```