---
title: "Project_Methods"
author: "Nathan Hogg"
date: "2026-04-15"
output: pdf_document
latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Define Data
Adam et al. "Clustering and superspreading potential of SARS-C-oV-2 infections in Hong Kong".
In:Nature Medicine(2020).doi:https://doi.org/10.1038/s41591-020-1092-0. 

```{r}
pairs <- read.csv("transmission_pairs.csv")

infector_counts <- table(pairs$infector.case)

all_infectees <- unique(pairs$infectee.case)
all_infectors <- unique(pairs$infector.case)
terminal_cases <- setdiff(all_infectees, all_infectors)

secondary_cases <- c(as.integer(infector_counts), rep(0, length(terminal_cases)))

freq_table <- as.data.frame(table(secondary_cases))
colnames(freq_table) <- c("secondary_cases", "frequency")

print(freq_table)

freq_table$secondary_cases <- as.integer(as.character(freq_table$secondary_cases))
raw_data <- rep(freq_table$secondary_cases, freq_table$frequency)


```

# ZIP Distributions
Application of Zero Inflated Poisson (ZIP) distribution to model the number of secondary cases per infector, accounting for excess zeros in the data.
## Results
Lambda (Poisson Mean): 1.398481 
Sigma (Zero-Inflation Probability): 0.5047316 
Log-Likelihood: -295.6031 
AIC: 595.2061 
Overall mean (lambda * (1 - sigma)): 0.6926233 
## Interpretation
The estimated Poisson Mean (Lambda) of approximately 1.398 suggests that, on average, an infector is expected to cause about 1.398 secondary cases in the absence of zero-inflation.
The Zero-Inflation Probability (Sigma) of approximately 0.505 indicates that about 50.5% of cases are "structural zeros", meaning they are not expected to cause any secondary cases
due to factors such as isolation, hospitalization, or other interventions.
The overall mean number serves as the effective reproduction number (R), which matches with Adam et al.'s estimate of R = 0.69. This suggests that, on average, each infector is expected
to cause less than one secondary case, indicating that the outbreak may be under control in this population.
The Akaike Information Criterium (AIC) will be used to compare the fit of the ZIP model to other models to determine which one best captures the distribution of secondary cases in the data.
Finally, we can check the prediction of the model by evaluating the predicted zeros against the observed zeros in the data:

\begin{align}
P(X=0) = \sigma + (1 - \sigma) \cdot e^{-\lambda} = 0.505 + 0.495 \cdot e^{-1.398} \approx 0.505 + 0.495 \cdot 0.247 = 0.505 + 0.122 = 0.627
\end{align}

As seen here, the model's prediction was correct within the observed data, which had a zero proportion of approximately 0.63. This suggests that the ZIP model is a good fit for the data, capturing the excess zeros effectively.

```{r}

# Log-Likelihood Function
zip_loglik <- function(params, data) {
  lambda <- params[1] # Poisson mean
  sigma <- params[2]  # Zero-inflation probability
  
  # Constraints: lambda > 0, 0 < sigma < 1
  if (lambda <= 0 || sigma <= 0 || sigma >= 1) return(-Inf)
  
  zeros <- data[data == 0]
  nonzeros <- data[data > 0]
  
  # Log-Likelihood for zeros: log(sigma + (1 - sigma) * exp(-lambda))
  ll_zeros <- length(zeros) * log(sigma + (1 - sigma) * exp(-lambda))
  
  # Log-Likelihood for non-zeros: log(1 - sigma) + log(Poisson probability)
  ll_nonzeros <- sum(log(1 - sigma) + dpois(nonzeros, lambda, log=TRUE))
  
  return(ll_zeros + ll_nonzeros)
}

# Negative log-likelihood for optim 
zip_negloglik <- function(params, data) {
  return(-zip_loglik(params, data))
}

# Fit the model
fit_zip <- function(data) {
  # Starting values
  p_zero <- mean(data == 0)
  start_sigma <- max(0.01, min(0.99, p_zero - 0.1))
  start_lambda <- mean(data) / (1 - start_sigma)
  
  result <- optim(
    par = c(start_lambda, start_sigma),
    fn = zip_negloglik,
    data = data,
    method = "L-BFGS-B",
    lower = c(1e-6, 1e-6),
    upper = c(Inf, 1 - 1e-6),
    hessian = TRUE
  )
  
  lambda_hat <- result$par[1]
  sigma_hat <- result$par[2]
  loglik <- -result$value
  aic <- 2 * 2 - 2 * loglik # 2 params
  
  cat("Lambda (Poisson Mean):", lambda_hat, "\n")
  cat("Sigma (Zero-Inflation Probability):", sigma_hat, "\n")
  cat("Log-Likelihood:", loglik, "\n")
  cat("AIC:", aic, "\n")
  cat("Overall mean (lambda * (1 - sigma)):", lambda_hat * (1 - sigma_hat), "\n")
  
  return(list(lambda = lambda_hat, sigma = sigma_hat, loglik = loglik, aic = aic, fit = result))
}

# Run on raw data
zip_result <- fit_zip(raw_data)

# Check predicted vs. observed zeros
n <- length(raw_data)
observed_zero_prop <- mean(raw_data == 0)

lambda <- zip_result$lambda
sigma <- zip_result$sigma

predicted_zero_prop <- sigma + (1 - sigma) * exp(-lambda)

cat("lambda:", lambda, "\n")
cat("sigma:", sigma, "\n")
cat("Observed Zero Proportion:", observed_zero_prop, "\n")
cat("Predicted Zero Proportion:", predicted_zero_prop, "\n")
cat("Observed zero count:", sum(raw_data == 0), "out of", n, "\n")
cat("Predicted zero count:", round(predicted_zero_prop * n), "out of", n, "\n")



```
# Negative Binomial Distribution

# Beta-Poisson Mixture Modeling
Application of the Beta-Poisson mixture model to secondary case data.
This model decomposes transmission into a Poisson-distributed contact process and a Beta-distributed infection probability, allowing for individual
level variation in infectiousness. The beta-Poison generalizes the negative binomial distribution and is the central
methodological contribution of Hilton & Hall.

## Results
Lambda (mean): 0.692623
Phi (overdispersion): 0.804689
Nu (inverse contacts): 0.0004845876
N (mean contacts): 2063.61
Log-likelihood: -276.2956
AIC: 558.5912\

## Interpretation
The mean number of secondary cases ($\lambda$) is fixed at the sample mean of approximately 0.693, consistent
with a subcritical outbreak where each case generates fewer than one secondary infection on average.

The overdispersion parameter ($\phi$) of approximately 0.805 governs the shape of the beta distribution from which individual
transmission probabilities are drawn. This corresponds to an implied negative binomial dispersion of
$\theta = \phi^{-1}=1.243$, indicating substantial individual-level heterogeneity in infectiousness, with some individuals being much more likely to transmit than others.

The inverse contact parameter ($N$) converged to approximately 0.0005, corresponding to $N \approx 2064$ mean contacts. this near-zero value of Nu indicates that the beta-Poisson
has collapsed to its negative binomial limit. In this limit ($N \to \infty$), the beta-distributed transmission probability scaled by $N$ converges to a gamma distribution, and the model
becomes equivalent to a negative binomial with mean $\lambda$ and dispersion $\theta = \phi^{-1}$:

\begin{align}
\lim_{N \to \infty} \text{BetaPoi}(\lambda, \Phi, N) = \text{NegBin}\left(\lambda, \theta = \frac{1}{\Phi}\right)
\end{align}
This result is consistent with the findings of Hilton and Hall (2024), who observed that the beta-Poisson converged to its negative binomial limit for the 
majority of their eight outbreak datasets. While the beta-Poisson achieved a substantially better fit than the ZIP model ($AIC=558.59$ vs. $595.21$), the improvement is driven
by the model's ability to capture over dispersion through its negative binomial limit rather than by the additional contact parameter $N$. 

The beta-Poisson's AIC of 558.59 includes a penalty for 3 parameters ($k = 3$; $\lambda$, $\phi$, $\nu$), compared to only 2 for the negative binomial. Since
the beta-Poisson has effectively reduced to the negative binomial here, we expect the negative binomial to achieve a similar log-likelihood with a lower AIC, making it the more parsimonious choice,
again consistent with Hilton and Hall's (2024) finding.

```{r}
# Beta-Poisson Mixture Model
bp_pmf <- function(x, lambda, phi, nu) {
  N <- 1 / nu
  alpha1 <- lambda * phi
  alpha2 <- (N - lambda) * phi
  
  if (alpha1 <= 0 || alpha2 <= 0 || N <= lambda) return(rep(-Inf, length(x)))
  
  sapply(x, function(k) {
    f <- function(p) dpois(k, lambda = p * N) * dbeta(p, alpha1, alpha2)
    val <- tryCatch(integrate(f, 0, 1, rel.tol = 1e-8)$value,
                    error = function(e) 0)
    if (val <= 0) return(-Inf)
    log(val)
  })
}

bp_negloglik <- function(params, data, lambda) {
  phi <- params[1]
  nu <- params[2]
  if (phi <= 0 || nu <= 0 || nu >= 1/lambda) return(1e10)
  ll <- sum(bp_pmf(data, lambda, phi, nu))
  if (is.na(ll) || is.infinite(ll)) return(1e10)
  return(-ll)
}

fit_bp <- function(data) {
  lambda <- mean(data)
  best <- list(value = 1e10)
  
  for (sp in c(0.1, 0.5, 1, 5)) {
    for (sn in c(0.01, 0.1, 0.5)) {
      sv <- sn / lambda
      if (sv >= 1/lambda) next
      
      res <- tryCatch(
        optim(c(sp, sv), bp_negloglik, data = data, lambda = lambda,
              method = "L-BFGS-B",
              lower = c(1e-6, 1e-8),
              upper = c(100, 1/lambda - 1e-6)),
        error = function(e) list(value = 1e10))
      
      if (res$value < best$value) best <- res
    }
  }
  
  phi_hat <- best$par[1]
  nu_hat <- best$par[2]
  N_hat <- 1 / nu_hat
  loglik <- -best$value
  aic <- 6 - 2 * loglik
  
  cat("Beta-Poisson MLE Results\n")
  cat("Lambda (mean):", lambda, "\n")
  cat("Phi (overdispersion):", phi_hat, "\n")
  cat("Nu (inverse contacts):", nu_hat, "\n")
  cat("N (mean contacts):", N_hat, "\n")
  cat("Log-likelihood:", loglik, "\n")
  cat("AIC:", aic, "\n")
  
  # Check if converged to negative binomial limit
  if (nu_hat < 0.01) {
    cat("\nNote: Nu ~ 0 (N → infty), beta-Poisson has converged to its\n")
    cat("negative binomial limit with theta =", 1/phi_hat, "\n")
  }
  
  return(list(lambda = lambda, phi = phi_hat, nu = nu_hat,
              N = N_hat, loglik = loglik, aic = aic))
}

#bp_result <- fit_bp(raw_data)

```

# Maximum Likelihood Estimation

# Branching Process




























