---
title: "Project_Methods"
author: "Nathan Hogg"
date: "2026-04-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Define Data
Adam et al. "Clustering and superspreading potential of SARS-C-oV-2 infections in Hong Kong".
In:Nature Medicine(2020).doi:https://doi.org/10.1038/s41591-020-1092-0. 

```{r}
pairs <- read.csv("transmission_pairs.csv")

infector_counts <- table(pairs$infector.case)

all_infectees <- unique(pairs$infectee.case)
all_infectors <- unique(pairs$infector.case)
terminal_cases <- setdiff(all_infectees, all_infectors)

secondary_cases <- c(as.integer(infector_counts), rep(0, length(terminal_cases)))

freq_table <- as.data.frame(table(secondary_cases))
colnames(freq_table) <- c("secondary_cases", "frequency")

print(freq_table)

raw_data <- rep(freq_table$secondary_cases, freq_table$frequency)
```

# ZIP Distributions
Application of Zero Inflated Poisson (ZIP) distribution to model the number of secondary cases per infector, accounting for excess zeros in the data.
```{r}

# Log-Likelihood Function
zip_loglik <- function(params, data) {
  lambda <- params[1] # Poisson mean
  sigma <- params[2]  # Zero-inflation probability
  
  # Constraints: lambda > 0, 0 < sigma < 1
  if (lambda <= 0 || sigma <- 0 || sigma >= 1) return(-Inf)
  
  zeros <- data[data == 0]
  nonzeros <- data[data > 0]
  
  # Log-Likelihood for zeros: log(sigma + (1 - sigma) * exp(-lambda))
  ll_zeros <- length(zeros) * log(sigma + (1 - sigma) * exp(-lambda))
  
  # Log-Likelihood for non-zeros: log(1 - sigma) + log(Poisson probability)
  ll_nonzeros <- sum(log(1 - sigma) + dpois(nonzeros, lambda, log=TRUE))
}


```
# Negative Binomial Distribution

# Beta-Poisson Mixture Modeling

# Maximum Likelihood Estimation

# Branching Process