---
title: "Project_Methods"
author: "Nathan Hogg"
date: "2026-04-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Define Data
Adam et al. "Clustering and superspreading potential of SARS-C-oV-2 infections in Hong Kong".
In:Nature Medicine(2020).doi:https://doi.org/10.1038/s41591-020-1092-0. 

```{r}
pairs <- read.csv("transmission_pairs.csv")

infector_counts <- table(pairs$infector.case)

all_infectees <- unique(pairs$infectee.case)
all_infectors <- unique(pairs$infector.case)
terminal_cases <- setdiff(all_infectees, all_infectors)

secondary_cases <- c(as.integer(infector_counts), rep(0, length(terminal_cases)))

freq_table <- as.data.frame(table(secondary_cases))
colnames(freq_table) <- c("secondary_cases", "frequency")

print(freq_table)

freq_table$secondary_cases <- as.integer(as.character(freq_table$secondary_cases))
raw_data <- rep(freq_table$secondary_cases, freq_table$frequency)


```

# ZIP Distributions
Application of Zero Inflated Poisson (ZIP) distribution to model the number of secondary cases per infector, accounting for excess zeros in the data.
## Results
Lambda (Poisson Mean): 1.398481 
Sigma (Zero-Inflation Probability): 0.5047316 
Log-Likelihood: -295.6031 
AIC: 595.2061 
Overall mean (lambda * (1 - sigma)): 0.6926233 
## Interpretation
The estimated Poisson Mean (Lambda) of approximately 1.398 suggests that, on average, an infector is expected to cause about 1.398 secondary cases in the absence of zero-inflation.
The Zero-Inflation Probability (Sigma) of approximately 0.505 indicates that about 50.5% of cases are "structural zeros", meaning they are not expected to cause any secondary cases
due to factors such as isolation, hospitalization, or other interventions.
The overall mean number serves as the effective reproduction number (R), which matches with Adam et al.'s estimate of R = 0.69. This suggests that, on average, each infector is expected
to cause less than one secondary case, indicating that the outbreak may be under control in this population.
The Akaike Information Criterium (AIC) will be used to compare the fit of the ZIP model to other models to determine which one best captures the distribution of secondary cases in the data.
Finally, we can check the prediction of the model by evaluating the predicted zeros against the observed zeros in the data:

\begin{align}
P(X=0) = \sigma + (1 - \sigma) \cdot e^{-\lambda} = 0.505 + 0.495 \cdot e^{-1.398} \approx 0.505 + 0.495 \cdot 0.247 = 0.505 + 0.122 = 0.627
\end{align}

As seen here, the model's prediction was correct within the observed data, which had a zero proportion of approximately 0.63. This suggests that the ZIP model is a good fit for the data, capturing the excess zeros effectively.

```{r}

# Log-Likelihood Function
zip_loglik <- function(params, data) {
  lambda <- params[1] # Poisson mean
  sigma <- params[2]  # Zero-inflation probability
  
  # Constraints: lambda > 0, 0 < sigma < 1
  if (lambda <= 0 || sigma <= 0 || sigma >= 1) return(-Inf)
  
  zeros <- data[data == 0]
  nonzeros <- data[data > 0]
  
  # Log-Likelihood for zeros: log(sigma + (1 - sigma) * exp(-lambda))
  ll_zeros <- length(zeros) * log(sigma + (1 - sigma) * exp(-lambda))
  
  # Log-Likelihood for non-zeros: log(1 - sigma) + log(Poisson probability)
  ll_nonzeros <- sum(log(1 - sigma) + dpois(nonzeros, lambda, log=TRUE))
  
  return(ll_zeros + ll_nonzeros)
}

# Negative log-likelihood for optim 
zip_negloglik <- function(params, data) {
  return(-zip_loglik(params, data))
}

# Fit the model
fit_zip <- function(data) {
  # Starting values
  p_zero <- mean(data == 0)
  start_sigma <- max(0.01, min(0.99, p_zero - 0.1))
  start_lambda <- mean(data) / (1 - start_sigma)
  
  result <- optim(
    par = c(start_lambda, start_sigma),
    fn = zip_negloglik,
    data = data,
    method = "L-BFGS-B",
    lower = c(1e-6, 1e-6),
    upper = c(Inf, 1 - 1e-6),
    hessian = TRUE
  )
  
  lambda_hat <- result$par[1]
  sigma_hat <- result$par[2]
  loglik <- -result$value
  aic <- 2 * 2 - 2 * loglik # 2 params
  
  cat("Lambda (Poisson Mean):", lambda_hat, "\n")
  cat("Sigma (Zero-Inflation Probability):", sigma_hat, "\n")
  cat("Log-Likelihood:", loglik, "\n")
  cat("AIC:", aic, "\n")
  cat("Overall mean (lambda * (1 - sigma)):", lambda_hat * (1 - sigma_hat), "\n")
  
  return(list(lambda = lambda_hat, sigma = sigma_hat, loglik = loglik, aic = aic, fit = result))
}

# Run on raw data
zip_result <- fit_zip(raw_data)

# Check predicted vs. observed zeros
n <- length(raw_data)
observed_zero_prop <- mean(raw_data == 0)

lambda <- zip_result$lambda
sigma <- zip_result$sigma

predicted_zero_prop <- sigma + (1 - sigma) * exp(-lambda)

cat("lambda:", lambda, "\n")
cat("sigma:", sigma, "\n")
cat("Observed Zero Proportion:", observed_zero_prop, "\n")
cat("Predicted Zero Proportion:", predicted_zero_prop, "\n")
cat("Observed zero count:", sum(raw_data == 0), "out of", n, "\n")
cat("Predicted zero count:", round(predicted_zero_prop * n), "out of", n, "\n")



```
# Negative Binomial Distribution

# Beta-Poisson Mixture Modeling

# Maximum Likelihood Estimation

# Branching Process